---
title: "Generating and calculating causal models"
author: "Cory Whitney"
date: "`r Sys.Date()`"
output: 
 html_document:
      toc: true
      toc_float: true
      toc_collapsed: false
bibliography: references/Causal_refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(causaleffect)
library(igraph)
```

# Causal maps

Make a causal map in `igraph` [@igraph]. For example the notation X -+ Z identifies and draws a directed edge (a causal arrow) from X to Z. These can be words or phrases but this does not look very good yet and we probably need to do a bit of work learning and 'hacking' igraph if we want these to be very verbose text boxes.

```{r test_fig}
test_fig <- graph.formula(W -+ X, W -+ Z, X -+ Z, Z -+ Y, X -+ Y, Y -+ X, simplify = FALSE)

plot(test_fig)
```

The fifth and sixth arrow of the "graph.formula" (connecting X and Y) are bidirected. We therefore have to add the following "set.edge.attribute" command:

```{r fig1 ammend}
test_fig_ammend <- set.edge.attribute(graph = test_fig, name = "description", index = c(5,6), value = "U")

plot(test_fig_ammend)
```

# Mathematical expressions

Get a mathematical expression for the 'interventional distribution' (do-Calculus) using the `causal.effect` function from the causaleffect library [@tikkaIdentifyingCausalEffects2017].

```{r ce1}
ce_formula <- causal.effect(y = "Y", x = "X", z = NULL, G = test_fig_ammend, expr = TRUE)

ce_formula
```

The result of this function is readable in Latex. We can replace the double slash '\\' with a single '\' so that it is also readable in Rmarkdown.

Or we can avoid that manual work and use the 'Concatenate and Print' `cat()` function. 

```{r cat ce1}
cat(ce_formula)
```

The results can be added to Rmd files when wrapped in '$' symbols in R markdown. This results in a LateX formula looks like this: 

$\sum_{W,Z}\left(\sum_{X}P(Y|W,X,Z)P(X|W)\right)P(Z|W,X)P(W)$

# Nutrition example

We generate a logical causal impact pathway for an intervention to planting trees for better nutrition [c.f. @whitneyIntegrationFruitTrees2018]. 

```{r message=FALSE, warning=FALSE}
#decision impact pathway

nutrition_decision <- graph.formula(PT -+  FY, #plant_trees -+ fruit_yield, 
                                    PT -+  C, #plant_trees -+ costs,
                                    FY -+  EF, #fruit_yield -+ eat_fruit,
                                    NK -+  EF, #nutrition_knowledge -+ eat_fruit,
                                    FY -+  S, #fruit_yield -+ sales,
                                    EF -+  N, #eat_fruit -+ nutrition,
                                    C -+  I, #costs -+ income,
                                    S -+  I, #sales -+ income,
                                    I -+  N, #income -+ nutrition,
                                    NN -+  N, #nutrition_needs -+ nutrition,
                   simplify = FALSE)

plot(nutrition_decision)

```

Use the `igraph` model to get the formula for the causal effect of the nutrition intervention to plant trees `PT` on nutrition outcomes `N`. Concatenate and print the formula output from the `causal.effect` output with `cat` from base R. 

```{r get_nutrition_formula, message=FALSE, warning=FALSE, results='hide'}
nutrition_formula <- causal.effect(y = "N", x = "PT", z = NULL, G = nutrition_decision, expr = TRUE)

cat(nutrition_formula)
```

$\sum_{NK,NN,FY,EF,S,I}P(N|PT,NK,NN,FY,EF,S,I)P(I|PT,FY,S)P(S|PT,FY)P(EF|PT,NK,FY)P(FY|PT)P(NN)P(NK)$

# References


